Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-15 latex/main.tex
Start on finalising mapping details

Input genome alignments are masked in sc2ts
so that deletions and ambiguous bases are treated as missing data by the HMM,
and once the samples are incorporated into the ARG,
these bases are imputed to carry the values inherited by the node in question.
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
and to incorporate them into the ARG we identified a set of [XXX] sites
most frequently involved in deletions [cite source?]
and post-hoc mapped alignment data for these sites to the ARG using parsimony.
Although sites are treated independently in this approach,
the approximation successfully identified the most frequent deletions,
such as the recurrent 21991-21993 (See Document~\ref{sec:supplement_dels}
for additional analysis).
The node representing the origin of Alpha, with 290,407 descendants,
is correctly associated with key lineage-defining deletions:
11288-11296 (ORF1A del3675-3677),
21765-21770 (Spike del69-70),
21991-21993 (Spike del144), and
28271 (non-coding).
The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021},
also emerges twice in the origin of Omicron -
in the node leading to the BA.1 variant (342,099 descendants),
and the node leading to the BA.4 variant (189,746 descendants).
To speed up inference and to avoid issues caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] ``problematic sites'' 
masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned to one of 2026 Pango lineages using pangolin (STAR Methods).
All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data 
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow for multiple sibling origination 
nodes (Document~\ref{sec:supplement_pango_lineages}).

@@

Furthermore, some of the Pango X lineages above originate from
shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent
events in the ARG (Figure~\ref{fig:subgraphs}B).
These Pango X lineages were designated around the same time.
These results highlight that the ARG can help clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

@@

[NOTE: putting this here for now, until we find the right place to slot it
in next draft.]
See Document~\ref{sec:supplement_pango_lineages} for extended analyses,
Figure~\ref{fig:pango_x_copying_patterns} for copying patterns,
and Document S2 for ARG visualations of all Pango X lineages.

@@

% This table is generated by the notebook tab_pango_x_events
\begin{table}
\caption{
ARG events associated with Pango X lineages.
Each row corresponds to the oldest node assigned the specified
Pango lineage and the ancestor of all nodes assigned that label
(except XBB, XBB.1 and XM; see STAR methods). [MORE DETAILS]
}
\label{tab:pango_x_lineages}
\begin{footnotesize}
\begin{tabular}{lrrrllrl}
\toprule
\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
pango & avt & left & right & left p & right p & samp & nested \\
\midrule
XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
XAL:3} \\
XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
\midrule
\multicolumn{5}{l}{Type II RE event}\\
alias & avt & left & right & left p & right p & desc & by pango \\
\midrule
Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 &
{BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
{XE:1116, BA.2:37, XH:2, XAF:1} \\
XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {BA.2:4, XBH:2} \\
XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
BF.3:2} \\ XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
XAF:1} \\
XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
BA.2:17} \\
\bottomrule
\multicolumn{5}{l}{Events closely derived from RE}\\
pango & alias & path len & time & mutations & samples \\
\midrule
XR & XQ & 3 & 11 & 4 & 17 \\
XU & XQ & 6 & 103 & 4 & 1 \\
XAM & XQ & 6 & 43 & 3 & 21 \\
XAA & XQ & 6 & 33 & 3 & 17 \\
XAG & XQ & 6 & 55 & 4 & 6 \\
XAP & Xx & 4 & 67 & 1 & 20 \\
XZ & Xx & 4 & 48 & 1 & 48 \\
XAE & Xx & 2 & 47 & 7 & 9 \\
XAF & XAF & 5 & 140 & 8 & 1 \\
XAL & XM & 3 & 66 & 2 & 3 \\
\midrule
\multicolumn{5}{l}{Non recombinant}\\
pango & muts & parent & samples & nested \\
\midrule
XP & 1 & \textbf{BA.1.1} & 45 & {} \\
XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
XBQ & 1 & CJ.1 & 14 & {} \\
XBK & 1 & CJ.1 & 7 & {} \\
XBE & 3 & BA.5.2 & 65 & {} \\
XB & 2 & B.1 & 192 & {} \\
XAZ & 3 & \textbf{BA.5} & 133 & {} \\
XAV & 3 & BA.5.1.24 & 13 & {} \\
XAU & 1 & BA.2 & 8 & {} \\
XAS & 1 & BA.4 & 77 & {} \\
XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table}

@@

\begin{table}
\caption{
Concordance among methods in characterizing Pango X lineages.
These comparisons focus on the Pango X lineages associated with
the Type 1 recombination events (Table~\ref{tab:pango_x_lineages})
For each method, we calculated the number of Pango X lineages
which have inferred parent Pango lineages and breakpoint intervals
concordant with those proposed by the community.
The denominators indicate the number of Pango X lineages
for which detection results are available from the original studies \cite{Alfonsi2024,Li2024CovRecomb}.
For the Pango X lineages with non-overlapping breakpoint intervals,
the distances between the inferred breakpoint intervals and 
those proposed by the community are shown (median and range).
}
\label{tab:method_concordance}
\begin{tabular}{lccc}
\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
\cmidrule(lr){2-2} \cmidrule(lr){3-4}
Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
\midrule
Sc2ts & 14 / 15 (93.3) & 13 / 15 (86.7) & 540.5 (315, 766)\\
RecombinHunt-GISAID & 13 / 13 (100.0) & 10 / 13 (76.9) & 764.0 (4, 1202)\\
RecombinHunt-Nextstrain & 12 / 14 (85.7) & 11 / 14 (78.6) & 1160.0 (559, 2785)\\
CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
\bottomrule
\end{tabular}
\end{table}

@@

Our analysis of the number of averted mutations reveals a spectrum of plausibility
across all the ``robust'' recombination events
(Figure~\ref{fig:averted_mutations}A).
Examples of highly plausible recombinants are XA and XBB
(for an extended analysis, see Document~\ref{sec:supplement_pango_lineages}).
The ``copying pattern'' of XA visually confirms
that a high number of recombination-informative sites support the suggested parents 
(Figure~\ref{fig:averted_mutations}D).
XW is an example of a marginally well-supported recombinant.
The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent.
% FIXME What's document S3?
The copying patterns of all the recombination events are shown in Document S3.

@@

\begin{itemize}
    \item The ARG and the sequence alignments have been deposited at Zenodo and are publicly available.
    \item All original code has been deposited in GitHub and is publicly available.
    \item Jupyter notebooks to perform the analyses in this paper are available on GitHub.
\end{itemize}

@@

\subsection*{Key resources table}

\textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}

@@

Sc2ts is implemented in Python building on core PyData infrastructure
such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
and Pandas ,
using pyfaidx\cite{Shirley2015efficient} to access alignment data
[TODO JK to fill this out]

@@

The SARS-CoV-2 ARGs inferred using sc2ts can be conveniently and efficiently analyzed
using the mature and feature-rich tskit library \cite{Kelleher2018efficient,Ralph2020}.
Tskit is widely used in population genetics, and has been incorporated into many downstream applications
(e.g., \cite{Haller2019,Speidel2019,TerasakiHart2021,Fan2022,Korfmann2022,Mahmoudi2022,Petr2022,Rasmussen2022,Zhang2023}).

@@

\subsubsection*{Data preprocessing}

In this study, we used the SARS-CoV-2 sequences reassembled in a recent study \cite{Hunt2024}
using an amplicon-aware assembly method (named ``Viridian'').
The Viridian sequences contain far fewer artifacts compared to their publicly deposited counterparts on GenBank.
Consequently, phylogenetic tree inference using the Viridian data led to more parsimonious phylogenies,
as seen by a large reduction in the number of likely artifactual reversions \cite{Hunt2024}.

We downloaded 4,484,157 Viridian sequences obtained
from samples collected from January, 2020 to March, 2023 (``batch 1'').
We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using MAFFT v7.525 \cite{Katoh2013},
with the option ‘--keeplength’ to retain the reference coordinate system, discarding insertions.
We converted the MAFFT alignments and all sample metadata to VCF Zarr format \cite{Czech2025},
which are available for download on Zenodo (URL).

Prior to ARG inference, we preprocessed the remaining sequences as follows.
We treated low-quality bases (represented by Ns), gaps and ambiguous characters in the sequences as missing data.
When encoded as missing data, the bases do not inform sample matching (see ``The Li and Stephens model'').

During inference, we filtered out the samples with low-quality sequences and low-precision sampling dates.
We kept the quality check-passing sequences used to infer the Viridian UShER phylogeny \cite{Hunt2024}.
We skipped low-quality sequences that had more than 500 missing sites
(except when building the ``base'' ARG, when the maximum number of allowed missing bases was set to 10,000 instead).
We also skipped $\sim$40,000 samples which had a collection date of December 31, 2020.
The sampling frequency of this date is alarmingly high compared to other dates (at most 15,000 samples per day).
We suspect that this batch of samples might be unusually enriched for wrong collection dates,
likely due to arbitrary assignment of the originally reported collection date of ``2020'' to be ``31-12-2020''.
Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
confirmed that some of the samples were not collected in 2020.

Sites which are enriched for systematic sequencing errors or frequently experience homoplasy
can cause artifactual signals of recombination (REF).
The bioinformatics community has identified and curated a set of problematic sites
that are recommended to be excluded before phylogenetic inference (REF).
Here we assembled a bespoke set of such problematic sites
using the information in the Viridian sequences.
To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
without any site masking.
Then, we identified the top 100 sites which had the highest mutation counts in the ARG (URL).
Only six of these sites overlapped with the ``Problematic Sites''
downloaded from the UCSC Genome Browser website (URL),
which did not have a high number of mutations in the preliminary ARG.
To infer the final ARG, we masked the problematic sites
in our custom set
\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}.

@@

\subsubsection*{Identification of robust recombination events}

[TODO This section needs a full pass. I've merged in a section that was in the
supplement with an extremely terse section on counting supporting loci.
Reference Figure~\ref{fig:recombinant_qc} for the scatter plot and 
labelled quadrants, and Figure~\ref{fig:copying_pattern_examples}
for examples of the corresponding copying patterns. Also reference
any additional examples that we include beyond A-D.]

% This para was *{Counting supporting loci for recombinant parents}

When counting such loci, we considered nearly adjacent sites (separated by at
most three bases) as single loci, to avoid over-counting support from
correlated loci, which would lead to artificial signals of recombination. Also,
when counting supporting loci for a focal parent, we penalized loci where the
character(s) in the recombinant matches that in the non-focal parent but not in
the focal parent.

Before characterizing recombination events in detail, we sought to remove
probable artifacts. Systematic errors in SARS-CoV-2 sequences associated with
sequencing protocols have been documented (Turakhia et al., 2022), which can
confound recombination detection (De Maio et al., 2024). By inspecting copying
patterns at sites that differ between recombinant parents, we find a convincing
signal when treating adjacent and near-adjacent sites as a single locus.
Artifactual recombination events can be characterized as having fewer than four
such loci supporting the left or right parent on either side of a suggested
breakpoint (STAR Methods). Of the 929 recombination events in the ARG, this
measure identifies 543 as artifactual, leaving 386 ``robust'' events (Figure
S1A). The copying patterns depicted in Figure S1B illustrate the genomic
evidence for three recombination events identified as artifactual, compared to
a marginally robust one.

Evidence that our simple filter effectively flags artifacts can be seen in the
clear association between the AmpliSeq V1 kit and recombination events that we
identify as artifactual. This primer scheme was used to sequence XX of the
samples associated with an artifactual recombination event, in contrast to 15
of the ``robust'' events (Figure 2A), strongly implicating it as a common source
of problematic data. Through further inspection, we observed that many
artifactual recombination events contain stretches of missing data
characteristic of amplicon dropouts (REF).

>>>>

Input genome alignments are masked in sc2ts
so that deletions and ambiguous bases are treated as missing data by the HMM,
and once the samples are incorporated into the ARG,
these bases are imputed to carry the values inherited by the node in question.
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
and to incorporate them into the ARG we identified a set of [XXX] sites
most frequently involved in deletions [cite source?]
and post-hoc mapped alignment data for these sites to the ARG using parsimony.
Although sites are treated independently in this approach,
the approximation successfully identified the most frequent deletions,
such as the recurrent 21991-21993 (See Document~\ref{sec:supplement_dels}
for additional analysis).
The node representing the origin of Alpha, with 290,407 descendants,
is correctly associated with key lineage-defining deletions:
11288-11296 (ORF1A del3675-3677),
21765-21770 (Spike del69-70),
21991-21993 (Spike del144), and
28271 (non-coding).
The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021},
also emerges twice in the origin of Omicron -
in the node leading to the BA.1 variant (342,099 descendants),
and the node leading to the BA.4 variant (189,746 descendants).
To speed up inference and to avoid issues caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] ``problematic sites'' 
masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned to one of 2058 Pango lineages using pangolin (STAR Methods).
All but 21,680 sampled sequences (0.87\%) agreed with the source sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data 
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058) have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow for multiple sibling origination 
nodes (Document~\ref{sec:supplement_pango_lineages}).

@@

Furthermore, some of the Pango X lineages above originate from
shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related lineages consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent
events in the ARG (Figure~\ref{fig:subgraphs}B).
These Pango X lineages were designated around the same time.
These results highlight that the ARG can help clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

@@

[NOTE: putting this here for now, until we find the right place to slot it
in next draft.]
See Document~\ref{sec:supplement_indepth_pango_x_lineages} for extended analyses,
Figure~\ref{fig:pango_x_copying_patterns} for copying patterns,
and Document S2 for ARG visualations of all Pango X lineages.

@@

% This table is generated by the notebook tab_pango_x_events
\begin{table}
\caption{
ARG events associated with Pango X lineages.
Each row corresponds to the oldest node assigned the specified
Pango lineage and the ancestor of all nodes assigned that label
(except XBB, XBB.1 and XM; see STAR methods). [MORE DETAILS]
}
\label{tab:pango_x_lineages}
\begin{footnotesize}
\begin{tabular}{lrrrllrl}
\toprule
\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
pango & avt & left & right & left p & right p & samp & nested \\
\midrule
XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
XAL:3} \\
XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {} \\
XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
\midrule
\multicolumn{5}{l}{Type II RE event}\\
alias & avt & left & right & left p & right p & desc & by pango \\
\midrule
Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 & {BA.2:156,
XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
{XE:1116, BA.2:37, XH:2, XAF:1} \\
XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
BF.3:2} \\
XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
XAF:1} \\
XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
BA.2:17} \\
\bottomrule
\multicolumn{5}{l}{Events closely derived from RE}\\
pango & alias & path len & time & mutations & samples \\
\midrule
XR & XQ & 3 & 11 & 4 & 17 \\
XU & XQ & 6 & 103 & 13 & 1 \\
XAM & XQ & 6 & 40 & 3 & 21 \\
XAA & XQ & 6 & 33 & 3 & 17 \\
XAG & XQ & 6 & 53 & 4 & 6 \\
XAP & Xx & 4 & 66 & 1 & 20 \\
XZ & Xx & 4 & 48 & 1 & 48 \\
XAE & Xx & 2 & 47 & 7 & 9 \\
XAF & XAF & 5 & 140 & 8 & 1 \\
XAL & XM & 3 & 65 & 2 & 3 \\
\midrule
\multicolumn{5}{l}{Non recombinant}\\
pango & muts & parent & samples & nested \\
\midrule
XP & 1 & \textbf{BA.1.1} & 45 & {} \\
XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
XBQ & 1 & CJ.1 & 14 & {} \\
XBK & 1 & CJ.1 & 7 & {} \\
XBE & 3 & BA.5.2 & 65 & {} \\
XB & 2 & B.1 & 192 & {} \\
XAZ & 3 & \textbf{BA.5} & 133 & {} \\
XAV & 3 & BA.5.1.24 & 13 & {} \\
XAU & 1 & BA.2 & 8 & {} \\
XAS & 1 & BA.4 & 77 & {} \\
XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table}

@@

\begin{table}
\caption{
Concordance among methods in characterizing Pango X lineages.
These comparisons focus on the Pango X lineages associated with
the Type 1 recombination events (Table~\ref{tab:pango_x_lineages})
For each method, we calculated the number of Pango X lineages
which have inferred parent Pango lineages and breakpoint intervals
concordant with those proposed by the community.
The denominators indicate the number of Pango X lineages
for which detection results are available from the original studies \cite{Alfonsi2024,Li2024CovRecomb}.
For the Pango X lineages with non-overlapping breakpoint intervals,
the distances between the inferred breakpoint intervals and 
those proposed by the community are shown (median and range).
}
\label{tab:method_concordance}
\begin{tabular}{lccc}
\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
\cmidrule(lr){2-2} \cmidrule(lr){3-4}
Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
\midrule
Sc2ts & 14 / 16 (87.5) & 14 / 16 (87.5) & 540.5 (315, 766)\\
RecombinHunt-GISAID & 13 / 15 (86.7) & 10 / 15 (66.7) & 764.0 (4, 1202)\\
RecombinHunt-Nextstrain & 12 / 15 (80.0) & 11 / 15 (73.3) & 1160.0 (559, 2785)\\
CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
\bottomrule
\end{tabular}
\end{table}

@@

Our analysis of the number of averted mutations reveals a spectrum of plausibility
across all the ``robust'' recombination events
(Figure~\ref{fig:averted_mutations}A).
Examples of highly plausible recombinants are XA and XBB
(for an extended analysis, see Document~\ref{sec:supplement_indepth_pango_x_lineages}).
The ``copying pattern'' of XA visually confirms
that a high number of recombination-informative sites support the suggested parents 
(Figure~\ref{fig:averted_mutations}D).
XW is an example of a marginally well-supported recombinant.
The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent.
% FIXME What's document S3?
The copying patterns of all the recombination events are shown in Document S3.

@@

\begin{itemize}
\item Snakemake workflow for aligning and converting the Viridian dataset to
VCF Zarr:
\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/viridian\_dataset/}
\end{itemize}

TODO add these to the list above with actual URLs
\begin{itemize}
    \item The ARG and the sequence alignments have been deposited at Zenodo and are publicly available.
    \item All original code has been deposited in GitHub and is publicly available.
    \item Jupyter notebooks to perform the analyses in this paper are available on GitHub.
\end{itemize}

@@

% \subsection*{Key resources table}

% \textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}

@@

Sc2ts is implemented in Python building on core PyData infrastructure
such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
and Pandas~\cite{Mckinney2010data} ,
using pyfaidx\cite{Shirley2015efficient} to access alignment data
and Numba~\cite{Lam2015numba} to accelerate certain alignment processing
tasks.
[TODO JK to fill this out]

@@

The SARS-CoV-2 ARGs inferred using sc2ts can be conveniently and efficiently analyzed
using the mature and feature-rich tskit library \cite{Kelleher2018efficient,Ralph2020}.
Tskit is widely used in population genetics, and has been incorporated into many downstream applications
(e.g., \cite{Haller2019,Speidel2019,TerasakiHart2021,Fan2022,Korfmann2023,Mahmoudi2022,Petr2023,Rasmussen2023,Zhang2023}).

@@

\subsubsection*{Pandemic scale alignment storage}
To provide the efficient access to alignment data required for inference with sc2ts, we
developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
Briefly, Zarr is a storage format for scientific data,
originally developed for the
\textit{Anopheles gambiae}
1000 Genomes Project~\cite{Anopheles2017genetic}, and
is now seeing widespread adoption across the 
sciences~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
Zarr is essentially a simple mechanism for storing n-dimensional array
data as a regular grid of compressed chunks, which provides 
high levels of data compression, efficient access across multiple
dimensions, and is ideally suited to modern cloud-based
deployments~\cite{Czech2025}.

We converted the Viridian v05 dataset\cite{Hunt2024}, consisting of 4,484,157 SARS-CoV-2
consensus sequences (125GiB over 48 FASTA files) and 30 metadata fields (1.4
GiB TSV file) to VCF Zarr using a 
reproducible Snakemake\cite{Koster2012snakemake} pipeline.
We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using 
MAFFT v7.475 \cite{Katoh2013},
with the option \texttt{--keeplength} to retain the reference coordinate system.
We then used \texttt{sc2ts import-alignments} and 
\texttt{sc2ts import-metadata} to convert all alignments and metadata to VCF
Zarr format, resulting in a single 401 MiB file 
(around 77X smaller than gzip compressed FASTA\cite{Czech2025}).

A classical tradeoff in bioinformatics is whether to store data 
in sample-major (FASTA) or variant-major (VCF) format, determining
whether the data can be accessed efficiently either by row or column
of the variant matrix. Because VCF Zarr is chunked in two dimensions,
we can now access either sample alignments or the variation
data for a given site across all samples efficiently.
While the VCF Zarr encoding of the MAFFT-aligned Viridian sequences
and metadata can be accessed via native Zarr libraries in several
popular programming languages (see Czech et al.\cite{Czech2025} 
for an overview), the sc2ts Python library provides a convenient interface. 
The \texttt{sc2ts.Dataset}
class provides efficient access to alignments for samples and 
variant data for sites along the genome, as well as methods to export
subsets of the data to FASTA if required.
Document~\ref{sec:supplement_sc2ts_usher_imputation} gives an example
of using this interface, where we retrieve alignment data for millions 
of samples in a few minutes on a standard desktop computer with minimal
memory overhead. See Czech et al.\cite{Czech2025} for more details 
and benchmarks.

Both the pipeline and converted Viridian dataset are available for download;
see Data and code availability for details.

\subsubsection*{Sample filtering and preprocessing}
The collection date for a sequence is an important piece of information
for sc2ts. We used the Viridian
consensus date (``Date\_tree``), based on information from COG-UK, GISAID, and
ENA/SRA as our source. Samples that have missing or incomplete dates (e.g., 2020-01)
were omitted, as were
the $\sim$40,000 samples which had a collection date of December 31, 2020.
As the sample count for this date is a major outlier for this period of
the pandemic, it seems likely that this 
batch is highly enriched for incorrect collection dates.
Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
confirmed that some of the samples were not collected in 2020.

The bioinformatics community has curated a set of problematic sites
for SARS-CoV-2 that are enriched for systematic sequencing errors or 
high levels of homoplasy~\cite{DeMaio2020issues}. 
Although these sites are recommended to be masked for UShER
inference~\cite{Turakhia2020},
the list has not been updated to take into account the systematic improvements 
made by the Viridian dataset, and we therefore assembled a custom set of sites to mask. 
To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
without any site masking.
Then, we identified the top 100 sites in terms of mutation counts in the ARG.
Only 11 of these sites 
(635, 8835, 11074, 11083, 15521, 16887, 21304, 21305, 21575, 21987 and 28253)
were present in the problematic sites file
(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}),
and most sites in this list were typical in terms of the numbers of mutations.
We therefore masked our 100 most-mutagenic sites to remove sequencing errors and improve
matching performance in the LS HMM.

Alignments read from the VCZ Zarr file were first preprocessed to mask any 
non nucleotide characters (A,C,G or T) as missing data. 
Then, we filtered 
any samples that had missing data at more than 500 sites, excluding
the 100 sites identified as problematic (this was 
set to 10,000 sites for the initial period up to 2020-03-01 where sampling
density is low).

@@

\subsubsection*{Post hoc parsimony mapping}
After the main inference step we mapped the variation data for 163 sites 
in the alignments to the ARG using the Fitch-Hartigan 
algorithm\cite{Fitch1971toward,Hartigan1973minimum} 
implemented in tskit. In contrast to the main inference process where 
deletion characters are masked as missing data, we allow for 
5 different states in these parsimony calculations: A,C,G,T and -.
By doing this, we can approximate the mutation events causing
deletions, but with the clear caveat that each site is treated 
independently. 
Using data provided by Li et al.\cite{Li2023indels} (Table S1) 
we chose 75 sites to remap

performed a large-scale analysis of indels
using a global GISAID dataset (sampled up to August, 2022).
We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
All of these indels are deletions and have high frequencies in the ARG
(sequence count divided by the total number of samples),
except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.

Based on the data of Table S4 of Li2023indels

\subsubsection*{Quality control of recombinants}

[TODO This section needs a full pass. I've merged in a section that was in the
supplement with an extremely terse section on counting supporting loci.
Reference Figure~\ref{fig:recombinant_qc} for the scatter plot and 
labeled quadrants, and Figure~\ref{fig:copying_pattern_examples}
for examples of the corresponding copying patterns. Also reference
any additional examples that we include beyond A-D.]

Before characterizing recombination events in detail,
we sought to remove probable artifactual recombinants
due to correlated bases
being treated as individual sites by the HMM.
Correlated bases can be caused by
systematic errors in genome sequences
associated with sequencing protocols \cite{Turakhia2020},
multi-nucleotide substitutions \cite{DeMaio2024},
genuine deletions, and
alignment errors (e.g., deletions due to MAFFT misalignment \cite{Hunt2024}).

We characterized artifactual recombination events
as having fewer than four such loci supporting the left or right parent
on either side of a suggested breakpoint.
We treated adjacent and near-adjacent sites
(separated by at most three bases) as single loci.
Also, when counting loci supporting a focal parent,
we penalized loci where the character(s)
in the recombinant matches that in the non-focal parent
but not in the focal parent.
